---
title: "MTF - exploring trends in beliefs over time"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

# Interesting questions

The *Monitoring The Future* survey has a lot of really interesting (and sometimes weird) questions that have been asked year after year, for decades. Some of the themes I'm curious about include:

**Attitudes about gender roles**

* Item Number: 07930, A: Men and women should be paid the same money if they do the same work 
* Item Number: 07940, B: Women should be considered as seriously as men for jobs as executives or politicians 
* Item Number: 07950, C: A woman should have exactly the same job opportunities as a man 
* Item Number: 07960, D: A woman should have exactly the same educational opportunities as a man 
* Item Number: 07970, E: It is usually better for everyone involved if the man is the achiever outside the home and * the woman takes care of the home and family 
* Item Number: 10470, A: One sees so few good or happy marriages that one questions it as a way of life 
* Item Number: 10480, B: It is usually a good idea for a couple to live together before getting married in order to find out whether they really get along 
* Item Number: 10490, C: Having a close intimate relationship with only one partner is too restrictive for the average person 
* Item Number: 10520, D: Being a father and raising children is one of the most fulfilling experiences a man can have 
* Item Number: 12170, E: Being a mother and raising children is one of the most fulfilling experiences a woman can have 
* Item Number: 10530, F: Most mothers should spend more time with their children than they do now 
* Item Number: 12180, G: Most fathers should spend more time with their children than they do now 

**Attitudes about gendered discrimination**

* Item Numbers: 12290-12350: ...Whether you think women are discriminated against in each of the following areas...
* Item Number: 10370, To what extent do you think the things listed below will prevent you from getting the kind of work you would like to have? B: Your sex

**Beliefs about social justice and institutions**

* Item Number: 01610, D: The way people vote has a major impact on how things are run in this country 
* Item Number: 01620, E: People who get together in citizen action groups to influence government policies can have a real effect 
* Item Number: 01520, L: [How important is it in your life] Working to correct social and economic inequalities
* Item Numers: 08380, 08390, 08430, 08440: How good or bad a job you feel each of the following organizations is doing for the country as a whole: Large corporations? Major labor unions? The national news media? The President and his administration? 
* Item Number: 12110, E: I get very upset when I see other people treated unfairly 

Even though similar or the same questions get asked each year, they have different variable names. I created mapping tables for the variables I'm interested in, so that it's easy to look up "what was X question's variable name in year Y?"

Some things I'm learning:

* `dplyr` would rather you not use rownames, and instead convert into a regular column of names-as-characters
* `mutate_at()` lets you mutate everything usng a function, and specifying `vars(-$COLNAME)` applies that function to everything *except* the specified column (without the `-`, it only mutates the specified columns)
* it's important to treat everything as characters rather than factors *<-- I wish I understood this more!*
* `pull()` takes a column from a tibble and returns a vector
* `:=` kets you use variables for column names (good when you're passing in those names from a list)
* `!!` expands an expression into a string (helpful to use in conjunction with the above or below)
* You can add names into the elements of a vector (maybe called [attributes](https://adv-r.hadley.nz/vectors-chap.html#attributes)?), which produces something like a dictionary. One way is to create your base vector `x`, and then assign names with `names(x) = c($CHARACTER VECTOR WITH NAMES, SAME LENGTH)`

```{r mapping_variables_by_year}
twelve_core_mapping = read.csv(file = "MappingTables/12core1991-2018.csv",
                               check.names = FALSE,
                               colClasses = c("character")
                               ) %>% 
  mutate_at(.,
            vars(-variable),
            .funs = toupper
            ) %>% 
  rename(.,
         new_variable_name = variable
         )

twelve_form3_mapping = read.csv(file = "MappingTables/12form3_1991-2018.csv", check.names = FALSE)
twelve_form6_mapping = read.csv(file = "MappingTables/12form6_1991-2018.csv", check.names = FALSE)
twelve_form7_mapping = read.csv(file = "MappingTables/12form7_1991-2018.csv", check.names = FALSE)

knitr::kable(head(twelve_core_mapping))
```

This function takes a variable name from an MTF SAS file, and returns a standardized (by me), more descriptive name.
```{r col_rename}
col_rename = function(old_name = old_name, year = year, mapping = mapping) {
  row_num = which(mapping[year] == old_name)
  
  mapping[row_num, 1]
}
```


```{r extract_demographics}
require(haven) # lets us read SAS files
require(stringr) # lets us subset a string using negative numbers

# Choices I made here, that could be different:
# A. use every variable from the mapping tibble
# B. so far, not validating input.
# C. I convert year to a character, and always use it as a character. This is helpful since it's a column name,
# ...but maybe not intuitve when reading the code (you expect year to be a number)

extract_demographics = function(path = path, years = years, mapping = mapping) {
  if (str_sub(path, -1) != "/") {
    path = str_c(path, "/")
  }
  
  # start with a blank tibble that we'll fill up by binding each year's resulting tibble
  demographics_all_years = tibble()
  
  for (year in years) {
    year = as.character(year)
    
    file_name = str_c(path, "y", year, "_1.sas7bdat")

    # variables to include in select() and rename() statements
    included_variable_old_names = pull(filter(mapping[year], !!sym(year) != "."))
    included_variable_new_names = map_chr(.x = included_variable_old_names,
                                          .f = col_rename,
                                          year = year,
                                          mapping = mapping
                                      )

    # NOTE: It's important that these two vectors retain ther order, so that the names are mapped correctly
    # ...it seems like that's what's happening, but I don't know R well enough to feel confident in this.
    rename_vector = included_variable_old_names
    names(rename_vector) = included_variable_new_names
    
    # variables that didn't exist for this given year, so we'll add them in as all NA using mutate() later
    excluded_variable_row_nums = which(mapping[year] == ".")
    excluded_variable_new_names = mapping[excluded_variable_row_nums, 1]
    
    # QUESTION: is there a less-code way to do everything above? maybe somehow getting a list of all variables names,
    # ...and then splitting on whether or not they're "."?

    # QUESTION: should year and grade be integers? factors? characters?
    
    # What's happening in these piped functions:
    # 1. Select all of the variables from our mapping table that existed for this year of the survey
    # 2. Add in (mutate) columns for year and grade
    # 3. rename varables from MTF names to standardized more descrptive names
    this_year_data = read_sas(data_file = file_name) %>% 
      select(.,
             one_of(included_variable_old_names)
             ) %>%
      mutate(.,
             year = as.integer(year),
             grade = 12
             ) %>%
      rename(.,
             !!rename_vector)
    
    ## NOTE: these for loops are ok, but I'm sure there's a way to do the same work using dplyr functions
    # ...like rename_at(), mutate_at(), across(), etc. I just can't figure it out :(
    
    # 4. add in any variables that were excluded that year as NA (but did exist in other years of interest),
    # ...so that the datasets can be stacked later
    for (excluded_var_new in excluded_variable_new_names) {
      this_year_data = mutate(this_year_data, !!excluded_var_new := NA)
    }

    # QUESTION: is there a better way to combine data frames vertically, by comlumn name??
    # need to sort alphabetically before binding rows
    this_year_data = this_year_data %>% select(sort(tidyselect::peek_vars()))
    
    # QUESTON: will I have trouble with factors?
    demographics_all_years = bind_rows(demographics_all_years, this_year_data)
  }
  
  demographics_all_years
}

t1 = extract_demographics(path = "~/Documents/Code/MTF/MTFData/12th_grade/",
                          years = 2013:2018,
                          mapping = twelve_core_mapping)
cat("Total number of rows:", nrow(t1))
knitr::kable(head(t1))
```


*not sure if I'll use this, but don't delete yet*
```{r extracting_relevant_data}
identifiers = c() # id, grade, year
demographics = c() # sex, region, SES (as parents' edu status)
questions = c() # list of questions
health_outcomes = c() # mental health? internalizing/externalizing behaviors?
```


